Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science

### 2025-09-11 conference-materials/2025_posit-conf/draft.qmd
Refactor conference slides for clarity and engagement
- Improved column layouts and margin handling for better screen distribution
- Applied color-coded emphasis to key headers and slide elements using CSS variables
- Enhanced narrative flow and pedagogy in behavioral science sections (e.g., added structured lists, clarified examples)
- Renamed "Telemetry Unraveled the User Journey" to "Telemetry Revealed the Journey" for clarity
- Added tabset-style walkthrough for "With vs Without Telemetry" workflows to guide different user paths
- Adjusted code block annotations to highlight step-by-step l

:::: {.columns}
::: {.column width="45%" .fragment}
<div class="tenor-gif-embed" data-postid="596341483704119392" data-share-method="host" data-aspect-ratio="1.32447" data-width="100%"><a href="https://tenor.com/view/monkey-computer-work-bug-developer-gif-596341483704119392">Monkey Computer GIF</a>from <a href="https://tenor.com/search/monkey-gifs">Monkey GIFs</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::

::: {.column width="45%" .fragment}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.32447" data-width="100%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
::::

@@

:::: {.columns}
::: {.column width="45%"}
### Raise hand if you've ever:
- Designed a UI with 10+ filters?
- Had a user ask for "one more dropdown"?
- Built a Shiny app with more `selectInput()` than insights?

::: {.fragment}
### Welcome! You're among friends 😊

*Think of our time together as a little therapy session for dropdown overload*
:::
:::

::: {.column width="45%" .fragment}
### 👋 Hi, I'm Jeremy!

**I'm a psychologist by training, but not the kind you're thinking of**

*Contrary to popular belief, most psychologists work with data. Not clients.*

::: {.fragment}
**PhD in Applied Psychology** → **Led Shiny teams at scale** → **Created {bidux} package**
:::
:::
::::

@@

::: {style="text-align: center; margin-top: 200px;"}
"Most dashboard failures aren't technical. They're psychological."
:::

@@

## Telemetry Unraveled the User Journey {background-color="#2c3e50"}

@@

::: {.fragment .text-xl style="text-align: center; margin-top: 10px;"}
**92% of users failed to find what they needed**
:::

@@

### What is it?

::: {.text-large}
The interdisciplinary study of how people **actually** make decisions (Spoiler: irrationally, predictably, consistently)
:::

:::: {.columns}
::: {.column width="50%"}
### Classic Example: The Paradox of Choice

::: {.text-large}
- [study details here]
:::
:::

::: {.column width="50%" .fragment}
### Study Results

![](img/jam-study.png){width="100%" style="border-radius: 8px; box-shadow: 0 4px 12px rgba(0,0,0,0.15);"}

::: {.fragment .text-xl style="color: var(--coral); font-weight: 600; margin-top: 30px;"}
**10x higher conversion with fewer options**
:::
:::
::::

@@

- **Cognitive Load**: Too many choices
- **Progressive Disclosure**: Reveal complexity gradually
- **Framing Effects**: Same data, different story
- **Choice Architecture**: Guide without restricting

::: {.notes}
- 20 seconds
- Now the paradox of choice is just one of many behavioral science concepts that directly apply to dashboards
- Taking this a step further...
- These aren't just theories...they're tools
:::

@@

:::: {.columns}
::: {.column width="30%"}
![](img/vw-bus.jpg){.r-stretch}
:::

::: {.column width="35%"}
### Think of it like a road trip

@@

::: {.column width="35%" .fragment}
### The BID Framework

@@

::: {.fragment}
![](img/bid-framework.png)
:::
:::
::::

@@

```{r}
library(bidux)

@@

::: {.fragment .text-xl style="text-align: center; margin-top: 30px;"}
**If you can't explain your dashboard's purpose in 10 words,<br>
users won't find answers in 100 filters**
:::

::: {.notes}
- 1 minute
- "Start with why - what question does this dashboard answer?"
- [CLICK] "10 word rule"
- Why this matters for your daily work: Clear purpose drives every technical decision you make
- Most devs think they do this, but telemetry says otherwise. We jump straight to technical implementation
- The cost: Every filter you add without purpose increases cognitive load exponentially
- This isn't just UX theory - it's engineering efficiency. When you clarify purpose first:
    1. You write less code because you know what to exclude
    2. Your data queries are more focused and performant
    3. User testing becomes objective rather than subjective
    4. You can defend design decisions with evidence, not opinions
- `bid_interpret` creates a structured object that feeds into later stages, making behavioral science actionable in your R workflow
- Think of it as writing better requirements that actually connect to user psychology
:::

@@

:::: {.columns}
::: {.column width="55%"}
```{r}
# Minimal input → Maximum insight
notice_stage <- bid_notice(
  previous_stage = interpret_stage,
  problem = "Users are struggling to find key insights",
  evidence = "Users report confusion and telemetry shows low engagement"
)

@@

:::: {.columns}
::: {.column width="50%"}
### Same Data, Different Story

@@

::: {.column width="50%"}
![Marketing: 🎉](img/ui-gain-frame.png){style="margin-bottom: 20px;"}

@@

```{r}
# Option A - with telemetry (SQLite or JSON)
issues <- bid_telemetry("dashboard_telemetry.sqlite")
count(issues, severity, sort = TRUE)

report <- bid_interpret(central_question = "What do users need first?") |>
  bid_notice_issue(issues[1, ]) |> # for use with telemetry
  bid_anticipate() |>
  bid_structure() |>
  bid_validate() |>
  bid_report(format = "html", include_diagrams = TRUE)

# Option B - without telemetry
report <- bid_interpret(central_question = "What do users need first?") |>
  bid_notice(
    problem = "Too many filters, unclear defaults",
    evidence = "User feedback and slow engagement"
  ) |>
  bid_anticipate() |>
  bid_structure() |>
  bid_validate() |>
  bid_report(format = "text", include_diagrams = FALSE)
```

::: {.notes}
- 1 minute
- "Complete workflow in one pipe"
- "Works with or without telemetry"
- bid_telemetry() provides a tidy tibble of issues compatible with dplyr.
- bid_notice_issue() is the bridge from a single issue to a BID Notice stage.
- The 5-stage chain (Interpret → Notice → Anticipate → Structure → Validate) culminates in bid_report(), which can produce text, HTML, or markdown.
- Output is a concise text report you can paste into PRs or tickets.
- Swap format = "html" if you want a styled artifact.
- No telemetry? Start at Interpret + Notice manually, then continue the same way.
- "Everything you need to fix your dashboard"
:::

>>>>

:::: {.columns}
::: {.column width="45%" .fragment}
<div class="tenor-gif-embed" data-postid="596341483704119392" data-share-method="host" data-aspect-ratio="1.32447" data-width="350px"><a href="https://tenor.com/view/monkey-computer-work-bug-developer-gif-596341483704119392">Monkey Computer GIF</a>from <a href="https://tenor.com/search/monkey-gifs">Monkey GIFs</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::

::: {.column width="45%" .fragment}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.32447" data-width="350px"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
::::

@@

:::: {.columns}
::: {.column width="50%"}
### Raise hand if you've ever: {style="color: var(--teal-light) !important;"}
- Designed a UI with 10+ filters?
- Had a user ask for "one more dropdown"?
- Built a Shiny app with more `selectInput()` than insights?
:::

::: {.column width="50%" .center}
### Welcome! You're among friends 😊 {style="color: var(--coral-light) !important;"}

*Think of our time together as a little therapy session for dropdown overload*
:::

::: {.column width="50%" style="margin-top: var(--space-8);" .fragment}
### 👋 Hi, I'm Jeremy! {style="color: var(--indigo-light) !important;"}

**I'm a psychologist by training, but probably not the kind you're thinking of**

*Contrary to popular belief, most psychologists work with data. Not clients.*
:::

::: {.column width="50%" style="margin-top: var(--space-8);" .fragment .center}
### My Journey {style="color: var(--indigo-light) !important;"}

**Applied Social Psychology** → <br>
**Full Stack Engineer** → <br>
**Led Shiny teams at scale** → <br>
**Created {bidux} package**
:::
::::

@@

::: {.insight-box style="margin-top: 300px !important;"}
"Most dashboard failures aren't technical. They're psychological."
:::

@@

## Telemetry Revealed the Journey {background-color="#2c3e50"}

```{r}
#| echo: false
#| fig-width: 14
#| fig-height: 7
#| fig.align: "center"

@@

::: {.fragment .text-xl style="text-align: center"}
**92% of users failed to find what they needed**
:::

@@

::: {.text-large}
**What is it?** The interdisciplinary study of how people *actually* make decisions<br>
(Spoiler: irrationally, yet predictably & consistently)
:::

::: { style="margin-top: var(--space-7);"}
:::: {.columns}
::: {.column width="45%" .fragment}
### Classic Example: The Paradox of Choice {style="color: var(--coral-light) !important;"}

::: {.text-large}
- **Researchers**: Sheena Iyengar & Mark Lepper
- **Setting**: Upscale grocery store sampling booth
- **Design**: 24 jams vs. 6 jams display
- **Key Finding**: More choice attracted browsers but paralyzed buyers
:::
:::

::: {.column width="45%" .fragment}
### Study Results {style="color: var(--teal-light) !important;"}

![](img/jam-study.png){width="100%" style="border-radius: 8px; box-shadow: 0 4px 12px rgba(0,0,0,0.15);"}
:::
::::
:::

::: {.fragment .center style="color: var(--coral); font-weight: 600; margin-top: 30px;"}
**10x higher conversion with fewer options**
:::

@@

:::: {.columns}
::: {.column width="45%"}
### Core Concepts {style="color: var(--coral-light) !important;"}

- **Cognitive Load**
    - Too many choices overwhelm users' mental capacity
- **Progressive Disclosure**
    - Reveal complexity gradually as users need it
- **Framing Effects**
    - Same data can tell completely different stories
- **Choice Architecture**
    - Guide decisions without restricting options
:::

::: {.column width="45%" .fragment}
### Dashboard Examples {style="color: var(--teal-light) !important;"}

- **Cognitive Load**
    - Start with 3 key filters, hide 15 advanced ones behind "More Options"
- **Progressive Disclosure**
    - Show summary → drill to details → reveal comparisons
- **Framing Effects**
    - "Revenue up 15%" vs "Revenue missed target by 5%"
- **Choice Architecture**
    - Smart defaults, logical grouping, visual hierarchy
:::
::::

::: {.notes}
- 30 seconds
- Now the paradox of choice is just one of many behavioral science concepts that directly apply to dashboards
- [CLICK] Let me show you specific examples of how these apply
- Cognitive Load: Instead of 18 filters overwhelming users, start with 3 essential ones and hide advanced options
- Progressive Disclosure: Show high-level KPIs first, then let users drill into details, then comparisons
- Framing Effects: The same revenue number can be "up 15%" or "missed target by 5%" - completely different user reactions
- Choice Architecture: Smart defaults, logical grouping, visual hierarchy guides users naturally
- These aren't just theories...they're actionable engineering principles
:::

@@

:::: {.columns}
::: {.column width="30%"}
![](img/vw-bus.jpg){.r-stretch}

### Think of it like a road trip

@@

::: {.column width="70%" .fragment}
### The BID Framework

@@

::: {.fragment}
![](img/bid-framework.png){style="box-shadow: none !important; width: 100%;"}
:::
:::
::::

@@

```{r}
#| code-line-numbers: "|3-5|6-10|11-17"
library(bidux)

@@

::: {.notes}
- 1 minute
- "Start with why - what question does this dashboard answer?"
- [CLICK] First highlight: central_question - this drives everything else
- [CLICK] Second highlight: data_story - this creates narrative structure
- [CLICK] Third highlight: user_personas - this grounds design in real needs
- Why this matters for your daily work: Clear purpose drives every technical decision you make
- Most devs think they do this, but telemetry says otherwise. We jump straight to technical implementation
- The cost: Every filter you add without purpose increases cognitive load exponentially
- This isn't just UX theory - it's engineering efficiency. When you clarify purpose first:
    1. You write less code because you know what to exclude
    2. Your data queries are more focused and performant
    3. User testing becomes objective rather than subjective
    4. You can defend design decisions with evidence, not opinions
- `bid_interpret` creates a structured object that feeds into later stages, making behavioral science actionable in your R workflow
- Think of it as writing better requirements that actually connect to user psychology
:::

@@

:::: {.columns}
::: {.column width="55%"}
```{r}
#| code-line-numbers: "|2-6|8"
# Minimal input → Maximum insight
notice_stage <- bid_notice(
  previous_stage = interpret_stage,
  problem = "Users are struggling to find key insights",
  evidence = "Users report confusion and telemetry shows low engagement"
)

@@

:::: {.columns}
::: {.column width="40%"}
### Same Data, Different Story

@@

::: {.column width="60%"}
![Marketing: 🎉](img/ui-gain-frame.png){style="margin-bottom: 20px;"}

@@

::: {.panel-tabset}

### With Telemetry

```{r}
#| code-line-numbers: "|1-2|4-8|10"
# Load and analyze existing usage data
issues <- bid_telemetry("dashboard_telemetry.sqlite")
count(issues, severity, sort = TRUE)

report <- bid_interpret(central_question = "What do users need first?") |>
  bid_notice_issue(issues[1, ]) |> # bridge from telemetry to BID
  bid_anticipate() |>
  bid_structure() |>
  bid_validate() |>
  bid_report(format = "html", include_diagrams = TRUE)
```

### Without Telemetry

```{r}
#| code-line-numbers: "|1-5|6-9|10"
report <- bid_interpret(central_question = "What do users need first?") |>
  bid_notice(
    problem = "Too many filters, unclear defaults",
    evidence = "User feedback and slow engagement"
  ) |>
  bid_anticipate() |>
  bid_structure() |>
  bid_validate() |>
  bid_report(format = "text", include_diagrams = FALSE)
```

::: {.notes}
- 1 minute
- "Complete workflow in one pipe - works with or without telemetry"
- [CLICK] With Telemetry tab first:
- [CLICK] First highlight: Load and analyze existing data
- [CLICK] Second highlight: The 5-stage BID pipeline
- [CLICK] Third highlight: Generate report
- [CLICK] Without Telemetry tab:
- [CLICK] First highlight: Manual problem definition
- [CLICK] Second highlight: Same 5-stage pipeline
- [CLICK] Third highlight: Text output for tickets/PRs
- bid_telemetry() provides a tidy tibble of issues compatible with dplyr
- bid_notice_issue() is the bridge from a single issue to a BID Notice stage
- The 5-stage chain culminates in bid_report(), which can produce text, HTML, or markdown
- Output is a concise report you can paste into PRs or tickets
- "Everything you need to fix your dashboard"
:::
